<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Quick Examples of Making Complex Heatmaps}
-->

Quick Examples of Making Complex Heatmaps
========================================

**Author**: Zuguang Gu ( z.gu@dkfz.de )

**Date**: `r Sys.Date()`

-------------------------------------------------------------

```{r global_settings, echo = FALSE, message = FALSE}
library(markdown)

library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 5,
    fig.height = 5)
options(markdown.HTML.stylesheet = "custom.css")

options(width = 100)
```

*There will be more examples in this vignette.*

## Combine pvclust and heatmap

**pvclust** package provides a robust way to test the stability of the clustering
by random sampling from original data. Here you can organize the heatmap by the clustering
returned from `pvclust()`.

```{r}
library(ComplexHeatmap)

library(MASS)
library(pvclust)
data(Boston)
boston.pv <- pvclust(Boston, nboot=100)
plot(boston.pv)
```

Since by default `pvclust` clusters columns by 'correlation' method, we scale columns for
`Boston` data set to see the relative trend.

```{r}
Boston_scaled = apply(Boston, 2, scale)
Heatmap(Boston_scaled, cluster_columns = boston.pv$hclust, heatmap_legend_title = "Boston")
```

## Change styles of legend

By default, for both discrete and continuous colors, the legend is represented as a discrete color bar.
But continuous colors can also be set to continuous color bar. You can change the legend style both
for heatmap legend and simple annotation legend.

```{r}
library(circlize)
mat = matrix(rnorm(100), 10)
ha = HeatmapAnnotation(df = data.frame(value = runif(10)), 
	col = list(value = colorRamp2(c(0, 1), c("white", "red"))),
	color_bar = "continuous")
Heatmap(mat, heatmap_legend_title = "mat", heatmap_legend_color_bar = "continuous", 
	top_annotation = ha)
```

## Make a same plot as heatmap()

```{r}
heatmap(mat, col = topo.colors(50))
```

Compare to the native `heatmap()`, `Heatmap()` can give more accurate interpolation
for colors for continous values.

```{r}
Heatmap(mat, col = topo.colors(50), row_hclust_width = unit(2, "cm"), 
    column_hclust_height = unit(2, "cm"))
```

## The measles vaccine heatmap

Following code reproduces the heatmap introduced [here](https://biomickwatson.wordpress.com/2015/04/09/recreating-a-famous-visualisation/) and [here](https://benjaminlmoore.wordpress.com/2015/04/09/recreating-the-vaccination-heatmaps-in-r/).

```{r, fig.width = 10, fig.height = 8}
mat = readRDS(paste0(system.file("extdata", package = "ComplexHeatmap"), "/measles.rds"))
ha1 = HeatmapAnnotation(dist1 = anno_barplot(colSums(mat), bar_width = 1, gp = gpar(col = NA, fill = "#FFE200"), 
    border = FALSE, axis = TRUE))
ha2 = rowAnnotation(dist2 = anno_barplot(rowSums(mat), bar_width = 1, gp = gpar(col = NA, fill = "#FFE200"), 
    border = FALSE, which = "row", axis = TRUE), width = unit(1, "cm"))
ha_column = HeatmapAnnotation(cn = function(index) {
    year = as.numeric(colnames(mat))
    which_decade = which(year %% 10 == 0)
    grid.text(year[which_decade], which_decade/length(year), 1, just = c("center", "top"))
})
Heatmap(mat, name = "cases", col = colorRamp2(c(0, 800, 1000, 127000), c("white", "cornflowerblue", "yellow", "red")),
    cluster_columns = FALSE, show_row_hclust = FALSE, rect_gp = gpar(col= "white"), show_column_names = FALSE,
    row_names_side = "left", column_title = 'Measles cases in US states 1930-2001\nVaccine introduced 1961',
    top_annotation = ha1, top_annotation_height = unit(1, "cm"),
    bottom_annotation = ha_column, bottom_annotation_height = grobHeight(textGrob("1900"))) + ha2

decorate_heatmap_body("cases", {
    i = which(colnames(mat) == "1961")
    x = i/ncol(mat)
    grid.lines(c(x, x), c(0, 1), gp = gpar(lwd = 2))
    grid.text("Vaccine introduced", x, unit(1, "npc") + unit(5, "mm"))
})
```