<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{5. Heatmap and Annotation Legends}
-->

Heatmap and Annotation Legends
========================================

**Author**: Zuguang Gu ( z.gu@dkfz.de )

**Date**: `r Sys.Date()`

-------------------------------------------------------------

```{r global_settings, echo = FALSE, message = FALSE}
library(markdown)
options(markdown.HTML.options = c(options('markdown.HTML.options')[[1]], "toc"))

library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 5,
    fig.height = 5)
options(markdown.HTML.stylesheet = "custom.css")

options(width = 100)
```

Legends for all heatmaps and row annotations are drawn together and legends for all column annotations are drawn together. 
The legends for heatmaps and legends for annotations are put in independent viewports.

```{r legend_default, fig.width = 8, fig.keep = "all"}
library(ComplexHeatmap)
library(circlize)

set.seed(123)
mat = matrix(rnorm(80, 2), 8, 10)
mat = rbind(mat, matrix(rnorm(40, -2), 4, 10))
rownames(mat) = paste0("R", 1:12)
colnames(mat) = paste0("C", 1:10)

ha_column = HeatmapAnnotation(df = data.frame(type1 = c(rep("a", 5), rep("b", 5))), 
    col = list(type1 = c("a" =  "red", "b" = "blue")))
ha_row = rowAnnotation(df = data.frame(type2 = c(rep("A", 6), rep("B", 6))), 
    col = list(type2 = c("A" =  "green", "B" = "orange")), width = unit(1, "cm"))

ht1 = Heatmap(mat, name = "ht1", column_title = "Heatmap 1", top_annotation = ha_column)
ht2 = Heatmap(mat, name = "ht2", column_title = "Heatmap 2")
ht_list = ht1 + ht2 + ha_row

draw(ht_list)
```

Side of legends can be set by `heatmap_legend_side` and `annotation_legend_side`.

```{r legend_side, fig.width = 8, fig.keep = "all"}
draw(ht_list, heatmap_legend_side = "left", annotation_legend_side = "bottom")
```

`show_heatmap_legend` and `show_annotation_legend` set visibility of legends.


```{r legend_show, fig.width = 8, fig.keep = "all"}
draw(ht_list, show_heatmap_legend = FALSE, show_annotation_legend = FALSE)
```

You can choose to only show some of the heatmap legends by setting `show_heatmap_legend` to 
a logical value when constructing single heatmaps.
Also `HeatmapAnnotation()` (or the shortcut function `columnAnnotation()` and `rowAnnotation()`) provides
`show_legend` argument to control visibility of annotation legends.

```{r legend_show_part, fig.width = 10}
ha_column = HeatmapAnnotation(df = data.frame(type1 = c(rep("a", 5), rep("b", 5))), 
    col = list(type1 = c("a" =  "red", "b" = "blue")), show_legend = FALSE)
ha_row = rowAnnotation(df = data.frame(type2 = c(rep("A", 6), rep("B", 6))), 
    col = list(type2 = c("A" =  "green", "B" = "orange")), show_legend = FALSE, width = unit(1, "cm"))

ht1 = Heatmap(mat, name = "ht1", column_title = "Heatmap 1", top_annotation = ha_column)
ht2 = Heatmap(mat, name = "ht2", column_title = "Heatmap 2", show_heatmap_legend = FALSE)
ht1 + ht2 + ha_row
```

**ComplexHeatmap** only generates legends for heatmaps and simple annotations. Self-defined legends
can be passed by `heatmap_legend_list` and `annotation_legend_list` as a list of `grob` objects.

**grid** package provides `legendGrob()` to construct a legend grob with certain style but
styles are still limited. For advanced users, they can construct a legend grob totally from ground by `frameGrob()`
and `packGrob()`.

```{r self_defined_annotation_legend, fig.width = 10}
ha = HeatmapAnnotation(points = anno_points(rnorm(10)))
ht2 = Heatmap(mat, name = "ht2", column_title = "Heatmap 2", top_annotation = ha, show_heatmap_legend = FALSE)
lgd = legendGrob(c("dots"), pch = 16)
draw(ht1 + ht2, annotation_legend_list = list(lgd))
```

Parameters for making the legend can be set by `heatmap_legend_param` (for heatmap) or `annotation_legend_param` (for annotations).
The parameters that can be set are as follows:

- `title`: title of the legend
- `title_gp`: graphic parameters for the legend title
- `color_bar`: style of the color bar, i.e. continuous or discrete
- `grid_height`: height of the small grid in the color bar, only works for discrete color bar
- `grid_width`: width of the color bar
- `grid_border`: border of the color bar
- `at`: break values shown on the legend
- `labels`: labels which correspond to the breaks values
- `labels_gp`: graphic parameters for legend labels

```{r heatmap_list_advanced, fig.width = 10}
df = data.frame(type = c(rep("a", 5), rep("b", 5)))
ha = HeatmapAnnotation(df = df, col = list(type = c("a" =  "red", "b" = "blue")),
    annotation_legend_param = list(type = list(title = "TYPE", title_gp = gpar(fontsize = 14), 
                                               labels_gp = gpar(fontsize = 8))))
ht1 = Heatmap(mat, name = "ht1", column_title = "Heatmap 1", top_annotation = ha)
ht2 = Heatmap(mat, name = "ht2", column_title = "Heatmap 2", 
    heatmap_legend_param = list(title = "Heatmap2", title_gp = gpar(fontsize = 8), labels_gp = gpar(fontsize = 14)))
ht1 + ht2
```

You can specify break values and break labels (both for character values and numeric values) by `at` and `labels` 
in corresponding `heatmap_legend_param` and `annotation_legend_param`.

```{r self_define_heatmap_legend, fig.width = 10}
ha = HeatmapAnnotation(df = df, col = list(type = c("a" =  "red", "b" = "blue")),
    annotation_legend_param = list(type = list(title = "TYPE", title_gp = gpar(fontsize = 14), 
        labels_gp = gpar(fontsize = 8), at = c("a", "b"), labels = c("A", "B"))))
ht1 = Heatmap(mat, name = "ht1", column_title = "Heatmap 1", top_annotation = ha,
    heatmap_legend_param = list(at = c(-3, 0, 3), labels = c("-three", "zero", "+three")))
ht1 + ht2
```

If you don't like the default discrete color bar for legends, you can specify `color_bar` to `continuous`.
For the simple annotation which contains continuous values, `color_bar` can also be set to `continuous`.

```{r}
ha = HeatmapAnnotation(df = data.frame(value = runif(10)), 
    col = list(value = colorRamp2(c(0, 1), c("white", "blue"))), 
    annotation_legend_param = list(color_bar = "continuous"))
Heatmap(mat, name = "ht1", top_annotation = ha, heatmap_legend_param = list(color_bar = "continuous"))
```

If you want to change default settings for all heatmaps/annotations, you can set it globally by `ht_global_opt()`.

```{r, fig.width = 10}
ht_global_opt(heatmap_legend_title_gp = gpar(fontsize = 16), annotation_legend_labels_gp = gpar(fontface = "italic"))
ha = HeatmapAnnotation(df = data.frame(value = runif(10)), 
    col = list(value = colorRamp2(c(0, 1), c("white", "blue"))), 
    annotation_legend_param = list(color_bar = "continuous"))
ht1 = Heatmap(mat, name = "ht1", column_title = "Heatmap 1", top_annotation = ha)
ht2 = Heatmap(mat, name = "ht2", column_title = "Heatmap 2", heatmap_legend_param = list(title_gp = gpar(fontsize = 8)))
ht1 + ht2
ht_global_opt(RESET = TRUE)
```

## Session info

```{r}
sessionInfo()
```
