<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Quick Examples of Making Complex Heatmaps}
-->

Quick Examples of Making Complex Heatmaps
========================================

**Author**: Zuguang Gu ( z.gu@dkfz.de )

**Date**: `r Sys.Date()`

-------------------------------------------------------------

```{r global_settings, echo = FALSE, message = FALSE}
library(markdown)

library(knitr)
knitr::opts_chunk$set(
    error = FALSE,
    tidy  = FALSE,
    message = FALSE,
    fig.align = "center",
    fig.width = 5,
    fig.height = 5)
options(markdown.HTML.stylesheet = "custom.css")

options(width = 100)
```


### Add more information for gene expression matrix

In gene expression matrix, rows correspond to genes. More information about genes can be attached after the expression
heatmap such as gene length and type of genes.

```{r expression_example, fig.width = 10, fig.height = 8}
library(ComplexHeatmap)

expr = readRDS(paste0(system.file(package = "ComplexHeatmap"), "/extdata/gene_expression.rds"))
mat = as.matrix(expr[, grep("cell", colnames(expr))])

type = gsub("s\\d+_", "", colnames(mat))
ha = HeatmapAnnotation(df = data.frame(type = type))

Heatmap(mat, name = "expression", km = 5, top_annotation = ha, 
    top_annotation_height = unit(4, "mm"), show_row_names = FALSE, 
    show_column_names = FALSE) +
Heatmap(expr$length, name = "length", col = colorRamp2(c(0, 100000), c("white", "orange")),
    width = unit(5, "mm")) +
Heatmap(expr$type, name = "type", width = unit(5, "mm")) +
Heatmap(expr$chr, name = "chr", col = rand_color(length(unique(expr$chr))), 
    width = unit(5, "mm"))
```

### Show number of alterations for OncoPrint

Following examples is <a href="http://www.cbioportal.org/faq.jsp#what-are-oncoprints">OncoPrint</a>.
The basic idea is to self define the heatmap body. Besides the default style which is 
provided by <a href="http://www.cbioportal.org/index.do">cBioPortal</a>, there are
additional barplots at both sides of the heatmap which show numbers of different alterations for
each sample and for each gene. Source code is available <a href="https://github.com/jokergoo/ComplexHeatmap/blob/master/vignettes/oncoprint.R">here</a>.

```{r, echo = FALSE, fig.width = 10, fig.height = 8}
source("oncoprint.R")
```

### Visualize genomic regions and other correspondance

Following examples visualizes correlation between methylation and expression, as well as other annotation information (data are randomly generated). In the heatmap, each row corresponds to a differentially methylated regions (DMRs). 
From left to right, heatmaps are:

1. methylation for each DMR (by rows) in samples.
2. direction of the methylation (one column heatmap), i.e. is methylation hyper in tumor or hypo?
3. expression for the genes that are associated with corresponding DMRs (e.g. closest gene).
4. significance for the correlation between methylation and expression (-log10(p-value)).
5. type of genes, i.e. is the gene a protein coding gene or a lincRNA?
6. annotation to gene models, i.e. is the DMR located in the intragenic region of the corresponding gene or the DMR is intergenic?
7. distance from the DMR to the TSS of the corresponding gene.
8. overlapping between DMRs and enhancers (Color shows how much the DMR is covered by the enhancers).

Source code is available <a href="https://github.com/jokergoo/ComplexHeatmap/blob/master/vignettes/genomic_regions.R">here</a>.

```{r, echo = FALSE, fig.width = 10, fig.height = 8}
library(circlize)
library(RColorBrewer)

lt = readRDS(paste0(system.file(package = "ComplexHeatmap"), "/extdata/meth.rds"))
list2env(lt, envir = environment())

ha = HeatmapAnnotation(df = data.frame(type = c(rep("Tumor", 10), rep("Control", 10))), 
    col = list(type = c("Tumor" = "red", "Control" = "blue")))
ha2 = HeatmapAnnotation(df = data.frame(type = c(rep("Tumor", 10), rep("Control", 10))), 
    col = list(type = c("Tumor" = "red", "Control" = "blue")), show_legend = FALSE)

# to make column order of expression matrix is the same as methylation matrix
column_tree = hclust(dist(t(meth)))

ht_list = 
    Heatmap(meth, name = "methylation", col = colorRamp2(c(0, 0.5, 1), c("blue", "white", "red")),
        cluster_columns = column_tree, top_annotation = ha, column_names_gp = gpar(fontsize = 8), km = 5, 
        column_title = "Methylation", column_title_gp = gpar(fontsize = 10), row_title_gp = gpar(fontsize = 10)) +
    Heatmap(direction, name = "direction", col = c("hyper" = "red", "hypo" = "blue"), 
        column_names_gp = gpar(fontsize = 8)) +
    Heatmap(expr[, column_tree$order], name = "expression", col = colorRamp2(c(-2, 0, 2), c("green", "white", "red")),
        cluster_columns = FALSE, top_annotation = ha2, column_names_gp = gpar(fontsize = 8), column_title = "Expression",
        column_title_gp = gpar(fontsize = 10)) +
    Heatmap(cor_pvalue, name = "-log10(cor_p)", col = colorRamp2(c(0, 2, 4), c("white", "white", "red")), 
        column_names_gp = gpar(fontsize = 8)) +
    Heatmap(gene_type, name = "gene type", col = brewer.pal(length(unique(gene_type)), "Set1"), 
        column_names_gp = gpar(fontsize = 8)) +
    Heatmap(anno, name = "anno_gene", col = brewer.pal(length(unique(anno)), "Set2"), 
        column_names_gp = gpar(fontsize = 8)) +
    Heatmap(dist, name = "dist_tss", col = colorRamp2(c(0, 10000), c("black", "white")), 
        column_names_gp = gpar(fontsize = 8)) +
    Heatmap(enhancer, name = "anno_enhancer", col = colorRamp2(c(0, 1), c("white", "orange")), 
        cluster_columns = FALSE, column_names_gp = gpar(fontsize = 8), column_title = "Enhancer", 
        column_title_gp = gpar(fontsize = 10))

ht_global_opt(heatmap_legend_title_gp = gpar(fontsize = 8, fontface = "bold"), 
              heatmap_legend_labels_gp = gpar(fontsize = 8))
draw(ht_list, newpage = FALSE, column_title = "Correspondence between methylation, expression and other genomic features", 
    column_title_gp = gpar(fontsize = 12, fontface = "bold"), heatmap_legend_side = "bottom")
ht_global_opt(RESET = TRUE)
```


## Combine pvclust and heatmap

**pvclust** package provides a robust way to test the stability of the clustering
by random sampling from original data. Here you can organize the heatmap by the clustering
returned from `pvclust()`.

```{r}
library(ComplexHeatmap)

library(MASS)
library(pvclust)
data(Boston)
boston.pv <- pvclust(Boston, nboot=100)
plot(boston.pv)
```

Since by default `pvclust` clusters columns by 'correlation' method, we scale columns for
`Boston` data set to see the relative trend.

```{r}
Boston_scaled = apply(Boston, 2, scale)
Heatmap(Boston_scaled, cluster_columns = boston.pv$hclust, heatmap_legend_param = list(title = "Boston"))
```

## Make a same plot as heatmap()

```{r}
mat = matrix(rnorm(100), 10)
heatmap(mat, col = topo.colors(50))
```

Compare to the native `heatmap()`, `Heatmap()` can give more accurate interpolation
for colors for continous values.

```{r}
Heatmap(mat, col = topo.colors(50), row_hclust_width = unit(2, "cm"), 
    column_hclust_height = unit(2, "cm"))
```

## The measles vaccine heatmap

Following code reproduces the heatmap introduced [here](https://biomickwatson.wordpress.com/2015/04/09/recreating-a-famous-visualisation/) and [here](https://benjaminlmoore.wordpress.com/2015/04/09/recreating-the-vaccination-heatmaps-in-r/).

```{r, fig.width = 10, fig.height = 8}
mat = readRDS(paste0(system.file("extdata", package = "ComplexHeatmap"), "/measles.rds"))
ha1 = HeatmapAnnotation(dist1 = anno_barplot(colSums(mat), bar_width = 1, gp = gpar(col = NA, fill = "#FFE200"), 
    border = FALSE, axis = TRUE))
ha2 = rowAnnotation(dist2 = anno_barplot(rowSums(mat), bar_width = 1, gp = gpar(col = NA, fill = "#FFE200"), 
    border = FALSE, which = "row", axis = TRUE), width = unit(1, "cm"))
ha_column = HeatmapAnnotation(cn = function(index) {
    year = as.numeric(colnames(mat))
    which_decade = which(year %% 10 == 0)
    grid.text(year[which_decade], which_decade/length(year), 1, just = c("center", "top"))
})
Heatmap(mat, name = "cases", col = colorRamp2(c(0, 800, 1000, 127000), c("white", "cornflowerblue", "yellow", "red")),
    cluster_columns = FALSE, show_row_hclust = FALSE, rect_gp = gpar(col= "white"), show_column_names = FALSE,
    row_names_side = "left", row_names_gp = gpar(fontsize = 10),
    column_title = 'Measles cases in US states 1930-2001\nVaccine introduced 1961',
    top_annotation = ha1, top_annotation_height = unit(1, "cm"),
    bottom_annotation = ha_column, bottom_annotation_height = grobHeight(textGrob("1900"))) + ha2

decorate_heatmap_body("cases", {
    i = which(colnames(mat) == "1961")
    x = i/ncol(mat)
    grid.lines(c(x, x), c(0, 1), gp = gpar(lwd = 2))
    grid.text("Vaccine introduced", x, unit(1, "npc") + unit(5, "mm"))
})
```

## What if my annotation name is too long?

